---
title: "WildCo_Data_Analysis_Intro"
author: "Oliver Fisher"
format: 
  html:
    self-contained: true
---

CRAN packages

```{r}
# A list of the required packages
list.of.packages <- c("activity",
                      "corrplot",
                      "cowplot",
                      "dplyr", 
                      "elevatr",
                      "gfcanalysis",  
                      "ggplot2",
                      "gridExtra",
                      "iNEXT",
                      "kableExtra",
                      "Hmsc",
                      "leaflet",
                      "lme4",
                      "lubridate",
                      "magrittr",
                      "MCMCvis",
                      "MODISTools",
                      "osmdata",
                      "pals",
                      "plotly",
                      "remotes",
                      "rmarkdown",
                      "sf",
                      "spOccupancy",
                      "stars",
                      "stringr",
                      "terra",
                      "tibble",
                      "tidyr", 
                      "unmarked",
                      "viridis",
                      "jtools",
                      "vegan",
                      "MuMIn",
                      "rgdal",
                      "usedist",
                      "taxize")

# A check to see which ones you have and which are missing
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

# Code which tells R to install the missing packages
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
```

Other Packages

```{r}
# We use the remotes package to access package repositories not on the 
# CRAN interface (e.g. on github)
library(remotes)

remotes::install_github("RS-eco/traitdata")
remotes::install_github("arcaravaggi/remBoot")
remotes::install_github("annam21/spaceNtime")
```

Google Earth Engine Package

```{r}
install.packages("rgee")
library("rgee")
```

Image rename code: unique code (e.g. placename_datetime) which will ensure that line of data you generate can be traced back to an image, regardless of how it is stored.

The Standard Form

-   *Project data* `proj.csv` a dataframe containing key information about the project itself, e.g. how the cameras were deployed and what the target features were.

-   *Image data* `img.csv` a dataframe containing all of the information contained within each image. This information is typically added by humans, but increasing we are using artificial intelligence to speed up this process.

-   *Deployment data* `dep.csv` a dataframe listing the activity of the camera traps involved in your study, and any issues encountered during deployments which may influence their analysis

-   *Camera data* `cam.csv` a dataframe all the cameras deployed in the project

```{r}
pro <- read.csv("data/raw_data/example_data/proj.csv", header=T)
img <- read.csv("data/raw_data/example_data/img.csv", header=T)
dep <- read.csv("data/raw_data/example_data/dep.csv", header=T)
cam <- read.csv("data/raw_data/example_data/cam.csv", header=T)

## Where data??
```


Foundations of detection
- The Independent detections dataf rame is the work horse of the vast majority of camera trap analyses,
- Effort data (names/location/date/time labeling)
- Observation by time intervals

Note Expand on section 7


Note see 7.4 for help with STAT313 project
